### 2023-02-21 tex/article.tex
Change citation commands

\begin{abstract}
Fronts affect phytoplankton growth and phenology by locally reducing stratification and increasing vertical nutrient supply.
Biomass peaks at fronts have been observed in-situ and linked to local nutrient upwelling, and reduced stratification over fronts has been shown to induce earlier blooms in numerical models.
However observation of these biophysical interactions through satellite imagery have been scarce, despite the opportunity to quantify them at synoptic scales.
Here we used twenty years of \as{sst} and al{chl} satellite data in a large region surrounding the Gulf Stream to quantify the impact of fronts on phytoplankton in contrasting regimes, from oligotrophy to bloom, and throughout the year.
We computed an heterogeneity index HI from \as{sst}, and used it to sort fronts into weak and strong fronts based on HI thresholds.
We observed that the localization of strong fronts corresponded to western boundary current fronts, and weak fronts to more ephemeral submesoscale fronts.
We compared al{chl} distributions over strong fronts, weak fronts and outside of fronts.
We assessed three metrics, the local enhancement of al{chl} over fronts, the global enhancement of \al{chl} due to fronts at the scale of the region, and the lag in spring bloom onset due to fronts.
We found that weak fronts lead to a local enhancement of al{chl} weaker than strong fronts, but because they are also more frequent they contribute equally to the regional \al{chl} budget.
We also find the the local enhancement of al{chl} was two to three times larger for the spring bloom than in the oligotrophic subtropical gyre.
We also provide observational evidence that blooms start earlier over fronts, by one to two weeks.
Nevertheless our results suggest that the spectacular impact of fronts at the local scale may be misleading, considering their impact on a regional scale budget remains limited.
\end{abstract}

@@

Phytoplankton form the basis of marine food webs and are key players in the ocean carbon cycle.
The transport of limiting nutrients to the sunlit euphotic layer by advective and convective processes, and the amount of light received by the cells --- which is closely related to the stratification of the water column --- are two important factors that control their growth.
As there are marked contrasts in nutrient and light availability in the ocean, it follows that the global ocean can be divided into different regional biomes (or bioregions), characterized by different phytoplankton abundances and seasonality \citep{longhurst_2007, vichi_2011a, bock_2022}.
The contrasts between biomes are largely explained by consistent physical forcings and environmental conditions, operating at the biome scale, which determine how the two main controlling factors, nutrient and light, limit growth.
For example, subtropical gyres are areas where wind-driven circulations induce a deepening of the thermocline and nutricline, resulting in oligotrophic biomes where productivity is relatively constant and low throughout the year; at higher latitudes, the nutricline is shallower and the strong seasonality of the vertical mixing will induce a multi-stage operation, with a time of reduced productivity and convective nutrient supply in winter when the mixing is strong, and a bloom in spring when the stratification sets in \citep{wilson_2005, williams_2011}.

In addition to these large-scale patterns, there has been considerable evidence over past years that submesoscale motions actively influence the nutrient and light environments at the horizontal scale of the order of 1--50km \citep[see reviews by][]{levy_2012, mahadevan_2016, levy_2018}.
Submesoscale motions arise dynamically through advective interactions involving mesoscale strain that continuously create sharp submesoscale density fronts, or at more steady wind-driven density fronts, such as western boundary currents \citep{thomas_2008, mcwilliams_2016, mahadevan_2020}.
These fronts are characterized by an energetic secondary vertical circulation, with upwelling on the warm side of the front, and downwelling on the cold side.
Submesoscale dynamics also involve restratification and suppression of vertical mixing at the front \citep{thomas_2008a}.
Thus submesoscale dynamics occurring at ocean fronts may affect phytoplankton in various ways.
Of interest here, the upward branch of the secondary circulation may enhance phytoplankton growth by transporting nutrients into the euphotic zone, while the downward branch may subduct biomass and excess nutrients into the subsurface \citep{calil_2011, omand_2015, hauschildt_2021}.
In addition, in highly seasonal regimes where productivity is slowed in winter due to deep mixing, submesoscale restratification may promote localized phytoplankton blooms before the large-scale outburst associated with seasonal stratification \citep{mahadevan_2012}.

Despite numerous local observations and a strong theoretical basis for these processes \citep[e.g. recent studies by][]{marrec_2018, little_2018, verneil_2019, ruiz_2019, uchida_2020, kessouri_2020, tzortzis_2021}, their integrated contribution at the scale of regional biomes is still largely unknown.
Indeed, as most fronts form, move and dissipate continuously on time scales of days to weeks, they are particularly difficult to sample, and this limitation is reinforced by the fact that only a limited number of fronts can be observed with in situ field observations.
Thus satellite-derived estimates of al{chl} (hereafter \as{chl}) although limited to the surface and an imperfect proxy for phytoplankton biomass, are the only data that allow to track the impact of fronts synoptically over large areas.
A first attempt to assess the contribution of fine scales to regional satellite \as{chl} budgets was based on a geostatistical analysis derived from data at 9km resolution \citep{doney_2003}, extended later in \citet{glover_2018}, with which they examined the change in spatial variance with distance.
This methodology was too coarse to reveal the impact of submesoscales but confirmed the important role of mesoscale eddies in stirring large scale gradients of phytoplankton abundance.
The role of submesoscales has been assessed with three different methods.
\citet{guo_2019} combined ocean color data with altimetry and drifting floats, and estimated that, over subtropical gyres of the global ocean, the respective contributions of mesoscales and submesoscales to high \as{chl} anomalies were comparable in magnitude.
\citet{keerthi_2022} proposed an approach based on deconvolution of local \as{chl} time series into different timescales; they observed that sub-seasonal time scales contributed roughly 30\% of the total satellite \as{chl} variance and were associated with small (\textless\ 100km) spacial scales.
Finally, the most quantitative approach was proposed by \citet{liu_2016}, which they applied to the North Pacific Subtropical Gyre.
They detected sea-surface temperature (\as{sst}) fronts by computing an index that measures the local heterogeneity of the \as{sst} field from satellite \as{sst} data.
This allowed them to compare satellite \as{chl} values over areas impacted by fronts (characterized by a large value of the heterogeneity index) with values over areas that were not impacted.
They found that the increase in \as{chl} over the fronts was negligible in summer but reached almost 40\% in winter.

Here we built on this last approach, and quantify the excess \as{chl} due to the presence of fronts at the scale of biomes.
This more global quantification will depend on several factors, since the local contribution of fronts to \as{chl} depends on many factors.
First it depends on how efficient fronts are at supplying nutrients, which itself depends on how deep the fronts reach into the nutricline, and of the seasonality of this supply.
It should be noted that the overall efficiency of fronts has been questioned \citep{levy_2018}; firstly because submesoscale fronts are more numerous in winter when convective nutrients input is also greatest; and secondly because the submesoscale vertical currents are often trapped within the mixed-layer, and may not reach the nutricline, which is often well below the base of the boundary mixed-layer.
Second, the overall contribution of fronts will differ between biomes, with submesoscale vertical advection of nutrients likely to be more important in oligotrophic biomes where other nutrient supply routes are scarce, and submesoscale restratification in blooming biomes.
Finally, the contribution of fronts will depend on their spatio-temporal footprint, which also varies seasonally \citep{callies_2015} and regionally \citep{mauzole_2022}.

Thus, more precisely, our intention is to explore and quantify how the contribution of fronts to biome-scale \as{chl} varies in three contrasted biomes, ranging from subtropical to subpolar, varies along the year, and varies with the occurrence and strength of fronts.
We focus our analysis on the North-Atlantic region surrounding the Gulf Stream, where multiple biomes and fronts of different strengths are found in a limited geographical area \citep{bock_2022} with strong seasonality.
In the south, our study area encompasses part of the North Atlantic subtropical gyre, characterized by an oligotrophic regime, year-long low productivity.
In the north, north of the Gulf Stream jet, is a more productive subpolar regime characterized by a recurrent spring bloom.
In between, there is a moderately productive regime, with maximum productivity in winter.
Another feature that makes this study area particularly relevant is that it has different types of fronts.
On the one hand, there are two strong persistent fronts, the Gulf Stream and the shelf-break front, which are held in place by topography and atmospheric circulation, and which are both associated with strong and deep-reaching vertical circulations \citep{flagg_2006, liao_2022}.
But there are also more ephemeral and weaker fronts, created by mesoscale strain that are continuously forming at more random locations \citep{drushka_2019, sanchez-rios_2020}.
We use satellite data of \as{chl} and \as{sst}, and extend on the approach of \citet{liu_2016}, to distinguish between persistent and ephemeral fronts.
We evaluate the impact of both types of fronts on \as{chl} on the basis of three indicators, the excess (or deficit) \as{chl} over fronts at the local scale of the front, the surplus \as{chl} attributable to fronts at the scale of regional biomes, and the change in the timing of the \as{chl} spring bloom over fronts.

@@

For \as{chl}, we used the L3 product distributed by ACRI-ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (SeaWIFS, MODIS Aqua \& Terra, MERIS, VIIRS-SNPP \& JPSS1, OLCI-S3A \& S3B) merged and reprocessed, available daily at 4km resolution \citep{chl}.

For \as{sst}, we used the European Space Agency Sea Surface Temperature Climate Change Initiative analysis product version 2.1 \citep{merchant_2019, good_2020, sst}, also available daily at 4km resolution over the period 2000--2020.
This product combines data from all available infrared sensors ((A)ATSR, SLSTR, and AVHRR sensors), ensuring good resolution where data are available, unlike other \as{sst} products which also include microwave and in-situ measurements, resulting in considerable smoothing of the \as{sst} field. %chktex 36
Where \as{sst} data is not available, spatial interpolation is performed to obtain a cloud free product which, at the cost of resolution on finer features, provides complete synoptic coverage of our large study area.
This interpolation tends to provide an underestimate of the detection of fronts, as the \as{sst} field is smoother over cloud-covered areas \citep{merchant_2019}.
However, the combination of several sensors allows to reduce these areas to a minimum.
Furthermore, we have only considered cloud-free pixels for our analysis, which ensures that cloudy areas are not taken into account in our quantification.

@@

The permanently oligotrophic regime to the south of our study area, characterized by warm waters and low \as{chl} (Fig.~\ref{fig:zone-separation}), is known as the subtropical gyre permanently stratified biome \citep{sarmiento_2004} or permanent deep \as{chl} maximum biome \citep{bock_2022}.
Their is no clear physical boundary to the northern limit of this permanent subtropical biome, so we have chosen the latitudinal limit of 32°N to delineate it, which roughly corresponds to the \(0.1 mg m^{-3}\) \as{chl} isocontour in annual mean \as{chl}.

North of 32°N, the seasonal subtropical biome is also mainly oligotrophic, with intermediate levels of \as{chl} and temperature, and characterized by slightly increased productivity in winter.
This second biome is known as the subtropical gyre seasonally stratified biome \citep{sarmiento_2004}.
It is bounded to the north by the meanders of the Gulf Stream jet.
The Gulf Stream jet conveys warm, salty waters poleward along the Florida coast up to Cape Hatteras (35°N), where the jet separates from the continental shelf and meanders essentially zonally.
The north wall of the jet, so called because of its steep temperature gradient, marks the sharp, sinuous and unsteady northern limit of this biome.

To the north of the Gulf Stream is the Slope sea which extends to the shelf-break, with colder and fresher waters \citep{linder_1998}.
Aligned with the shelf break, a persistent front with an intensified surface jet separates the shelf waters (excluded from this study) from the slope sea.
This highly productive subpolar biome, known as subpolar waters \citep{sarmiento_2004} and high-chlorophyll-bloom \citep{bock_2022}, is characterized by a strong spring bloom whose onset is tied to the spring stratification of the mixed-layer.

@@

Our front detection method aims to sort the domain into pixels that are located (or not) over fronts, at each daily time step.
We follow the approach of \citet{liu_2016} that builds upon the well established Cayula-Cornillon algorithm \citep{cayula_1992, belkin_2009}, and identify pixels as belonging to fronts when the region in the vicinity of the pixel is characterized by high \as{sst} heterogeneity.
More precisely, on each pixel and for each time step, we compute an Heterogeneity Index (HI) which quantifies the heterogeneity of the \as{sst} in a square window comprising a few pixels (\(7 \times 7\)) centered on the pixel of interest.
Thus HI measures the \as{sst} heterogeneity at spatial scales less than 30km.
The HI is defined as the weighted sum of the skewness \(\gamma\), standard deviation \(\sigma\), and bimodality \(B\) of \as{sst} within the window (\(\mathrm{HI} = a \left( b \gamma + c \sigma + d B \right)\), with a, b, c, d constant normalization coefficients).

We have adapted the original formulation of \citet{liu_2016} with minor modification in the computation of bimodality and in the relative weighing of each component which are described below.
We computed the bimodality as the L2 norm of the difference between the \as{sst} histogram (with bins of 0.1°C) and a Gaussian fit of the histogram.
We chose this method because it was simpler to implement and more robust than that of \citet{liu_2016} which computed the normalized absolute difference between a polynomial fit of order 5 of the \as{sst} histogram and a Gaussian fit of the histogram, the polynomial fit often being poorly constrained.
We normalized each component by its variance (b, c and d being defined as the inverse of the standard deviation of each component computed over one year), as opposed to \citet{liu_2016} who normalized the coefficients by their annual maximum.
This avoided putting too much weight on extrema.
Finally we normalized HI (coefficient a) such that 95\% of values are below an arbitrary value of 9.5.
For simplicity, the normalization coefficients were computed for year 2007 and used over the entire time series.
We performed sensitivity tests on the different parameters used to compute HI, namely the number of pixels in the rolling window and the choice of normalization coefficients.
An example of the resulting HI over a single front is given on Fig.~\ref{fig:zoom}; elevated HI values are located along the curved \as{sst} gradient to the north east of the window.

At each time step, we sorted the pixels into those belonging to strong fronts (defined as \(\mathrm{HI} > 10\)), those belonging to weak fronts (defined as \(5 < \mathrm{HI} < 10\)), and those that do not belong to fronts (when \(\mathrm{HI} < 5\), and called them ``background'' in the following).
Sorting fronts by range of HI enabled us to roughly separate the quasi-permanent fronts, which are associated with the strongest \as{sst} gradients (strong fronts), from the more ephemeral ones, associated with weaker gradients (weak fronts).
The choice of the two HI threshold is somehow arbitrary, but it is supported by the HI distributions presented below.

\begin{figure*}
  \includegraphics[width=12cm]{front_example.pdf}
  \caption{
    The \as{sst}, \as{chl}, and heterogeneity index (HI) of a front on the 7 July 2007.
    The plain and dashed contours correspond to HI values of 5 and 10.
    This front is categorized as weak.
    \as{chl} are elevated inside the front.
  }%
  \label{fig:zoom}
\end{figure*}

@@

Our definition of weak fronts and strong fronts is primarily based on thresholds, derived from the HI distributions (Fig.~\ref{fig:chl-vs-hi}, red dashed line).
In each biome, the majority of HI values are below 5, and as mentioned before, we used this threshold to distinguish pixels in the background from those over fronts (when \(\mathrm{HI} > 5\)).
Secondly, the number of points in the HI distribution decreases sharply as the HI value increases above 5, reflecting the fact that fronts with stronger \as{sst} gradients are much less frequent than fronts with weaker gradients, as one would expect.
We used the HI threshold of 10 to distinguish weak fronts from strong fronts.
This separation is imperfect, as seen for instance in Fig.~\ref{fig:zoom} where a few pixels with HI values larger than 10 appear at the core of an otherwise weak front.

\begin{figure*}
  \includegraphics[width=12cm]{chl_vs_hi.pdf}
  \caption{
    Normalized distribution of the Heterogeneity Index (HI, red dashed line) within each biome, and distribution of \as{chl} as a function of HI (representing front strength), over the full period 2000--2020.
    Shown are the median value of the \as{chl} distributions (solid black line), and 1st and 3rd quartiles (dashed lines).
    Note that 0.5\% of pixels have outstanding large HI values and are not included here.
  }%
  \label{fig:chl-vs-hi}
\end{figure*}

However, the choice of these two HI thresholds is also guided by the resulting global spatial climatology of weak and strong fronts (Fig.~\ref{fig:frt-occurrence}).
Weak fronts are abundant, and more or less evenly distributed, over a broad band around and north of the Gulf Stream jet (Fig.~\ref{fig:frt-occurrence}a).
To the south of the Gulf Stream jet, weak fronts are less present, with nevertheless more fronts on the edges of the subtropical gyre (around 28°N) than in its center.
This distribution of weak fronts is consistent with the predominance of mesoscale variability observed along the Gulf Stream system from satellite altimetry \citep{zhai_2008}, and the injection of eddy kinetic energy north of the Gulf Stream jet by the Gulf Stream extension.
It is thus consistent with the generation process of submesoscale fronts through mesoscale strain.

@@

With the chosen thresholds, the areal proportion of weak fronts tends to increase from South to North, with 7\%, 19\% and 42\% of HI values comprised between 5 and 10 in the permanent subtropical, seasonal subtropical, and subpolar biome, respectively.
Regarding strong fronts, there are only present in the seasonal subtropical and subpolar biome (as expected from the fact that the Gulf Stream and shelf-break fronts are located within these biomes), where HI values above 10 account for 6\% and 17\%, respectively.

>>>>

\begin{abstract}
Fronts affect phytoplankton growth and phenology by locally reducing stratification and increasing vertical nutrient supply.
Biomass peaks at fronts have been observed in-situ and linked to local nutrient upwelling, and reduced stratification over fronts has been shown to induce earlier blooms in numerical models.
However observation of these biophysical interactions through satellite imagery have been scarce, despite the opportunity to quantify them at synoptic scales.
Here we used twenty years of \as{sst} and al{chl} satellite data in a large region surrounding the Gulf Stream to quantify the impact of fronts on phytoplankton in contrasting regimes, from oligotrophy to bloom, and throughout the year.
We computed an heterogeneity index \as{hi} from \as{sst}, and used it to sort fronts into weak and strong fronts based on \as{hi} thresholds.
We observed that the localization of strong fronts corresponded to western boundary current fronts, and weak fronts to more ephemeral submesoscale fronts.
We compared al{chl} distributions over strong fronts, weak fronts and outside of fronts.
We assessed three metrics, the local enhancement of al{chl} over fronts, the global enhancement of \al{chl} due to fronts at the scale of the region, and the lag in spring bloom onset due to fronts.
We found that weak fronts lead to a local enhancement of al{chl} weaker than strong fronts, but because they are also more frequent they contribute equally to the regional \al{chl} budget.
We also find the the local enhancement of al{chl} was two to three times larger for the spring bloom than in the oligotrophic subtropical gyre.
We also provide observational evidence that blooms start earlier over fronts, by one to two weeks.
Nevertheless our results suggest that the spectacular impact of fronts at the local scale may be misleading, considering their impact on a regional scale budget remains limited.
\end{abstract}

@@

Phytoplankton form the basis of marine food webs and are key players in the ocean carbon cycle.
The transport of limiting nutrients to the sunlit euphotic layer by advective and convective processes, and the amount of light received by the cells --- which is closely related to the stratification of the water column --- are two important factors that control their growth.
As there are marked contrasts in nutrient and light availability in the ocean, it follows that the global ocean can be divided into different regional biomes (or bioregions), characterized by different phytoplankton abundances and seasonality \parencite{longhurst_2007, vichi_2011a, bock_2022}.
The contrasts between biomes are largely explained by consistent physical forcings and environmental conditions, operating at the biome scale, which determine how the two main controlling factors, nutrient and light, limit growth.
For example, subtropical gyres are areas where wind-driven circulations induce a deepening of the thermocline and nutricline, resulting in oligotrophic biomes where productivity is relatively constant and low throughout the year; at higher latitudes, the nutricline is shallower and the strong seasonality of the vertical mixing will induce a multi-stage operation, with a time of reduced productivity and convective nutrient supply in winter when the mixing is strong, and a bloom in spring when the stratification sets in \parencite{wilson_2005, williams_2011}.

In addition to these large-scale patterns, there has been considerable evidence over past years that submesoscale motions actively influence the nutrient and light environments at the horizontal scale of the order of 1--50km \parencite[see reviews by][]{levy_2012, mahadevan_2016, levy_2018}.
Submesoscale motions arise dynamically through advective interactions involving mesoscale strain that continuously create sharp submesoscale density fronts, or at more steady wind-driven density fronts, such as western boundary currents \parencite{thomas_2008, mcwilliams_2016, mahadevan_2020}.
These fronts are characterized by an energetic secondary vertical circulation, with upwelling on the warm side of the front, and downwelling on the cold side.
Submesoscale dynamics also involve restratification and suppression of vertical mixing at the front \parencite{thomas_2008a}.
Thus submesoscale dynamics occurring at ocean fronts may affect phytoplankton in various ways.
Of interest here, the upward branch of the secondary circulation may enhance phytoplankton growth by transporting nutrients into the euphotic zone, while the downward branch may subduct biomass and excess nutrients into the subsurface \parencite{calil_2011, omand_2015, hauschildt_2021}.
In addition, in highly seasonal regimes where productivity is slowed in winter due to deep mixing, submesoscale restratification may promote localized phytoplankton blooms before the large-scale outburst associated with seasonal stratification \parencite{mahadevan_2012}.

Despite numerous local observations and a strong theoretical basis for these processes \parencite[e.g. recent studies by][]{marrec_2018, little_2018, verneil_2019, ruiz_2019, uchida_2020, kessouri_2020, tzortzis_2021}, their integrated contribution at the scale of regional biomes is still largely unknown.
Indeed, as most fronts form, move and dissipate continuously on time scales of days to weeks, they are particularly difficult to sample, and this limitation is reinforced by the fact that only a limited number of fronts can be observed with in situ field observations.
Thus satellite-derived estimates of al{chl} (hereafter \as{chl}) although limited to the surface and an imperfect proxy for phytoplankton biomass, are the only data that allow to track the impact of fronts synoptically over large areas.
A first attempt to assess the contribution of fine scales to regional satellite \as{chl} budgets was based on a geostatistical analysis derived from data at 9km resolution \parencite{doney_2003}, extended later in \textcite{glover_2018}, with which they examined the change in spatial variance with distance.
This methodology was too coarse to reveal the impact of submesoscales but confirmed the important role of mesoscale eddies in stirring large scale gradients of phytoplankton abundance.
The role of submesoscales has been assessed with three different methods.
\textcite{guo_2019} combined ocean color data with altimetry and drifting floats, and estimated that, over subtropical gyres of the global ocean, the respective contributions of mesoscales and submesoscales to high \as{chl} anomalies were comparable in magnitude.
\textcite{keerthi_2022} proposed an approach based on deconvolution of local \as{chl} time series into different timescales; they observed that sub-seasonal time scales contributed roughly 30\% of the total satellite \as{chl} variance and were associated with small (\textless\ 100km) spacial scales.
Finally, the most quantitative approach was proposed by \textcite{liu_2016}, which they applied to the North Pacific Subtropical Gyre.
They detected sea-surface temperature (\as{sst}) fronts by computing an index that measures the local heterogeneity of the \as{sst} field from satellite \as{sst} data.
This allowed them to compare satellite \as{chl} values over areas impacted by fronts (characterized by a large value of the heterogeneity index) with values over areas that were not impacted.
They found that the increase in \as{chl} over the fronts was negligible in summer but reached almost 40\% in winter.

Here we built on this last approach, and quantify the excess \as{chl} due to the presence of fronts at the scale of biomes.
This more global quantification will depend on several factors, since the local contribution of fronts to \as{chl} depends on many factors.
First it depends on how efficient fronts are at supplying nutrients, which itself depends on how deep the fronts reach into the nutricline, and of the seasonality of this supply.
It should be noted that the overall efficiency of fronts has been questioned \parencite{levy_2018}; firstly because submesoscale fronts are more numerous in winter when convective nutrients input is also greatest; and secondly because the submesoscale vertical currents are often trapped within the mixed-layer, and may not reach the nutricline, which is often well below the base of the boundary mixed-layer.
Second, the overall contribution of fronts will differ between biomes, with submesoscale vertical advection of nutrients likely to be more important in oligotrophic biomes where other nutrient supply routes are scarce, and submesoscale restratification in blooming biomes.
Finally, the contribution of fronts will depend on their spatio-temporal footprint, which also varies seasonally \parencite{callies_2015} and regionally \citep{mauzole_2022}.

Thus, more precisely, our intention is to explore and quantify how the contribution of fronts to biome-scale \as{chl} varies in three contrasted biomes, ranging from subtropical to subpolar, varies along the year, and varies with the occurrence and strength of fronts.
We focus our analysis on the North-Atlantic region surrounding the Gulf Stream, where multiple biomes and fronts of different strengths are found in a limited geographical area \parencite{bock_2022} with strong seasonality.
In the south, our study area encompasses part of the North Atlantic subtropical gyre, characterized by an oligotrophic regime, year-long low productivity.
In the north, north of the Gulf Stream jet, is a more productive subpolar regime characterized by a recurrent spring bloom.
In between, there is a moderately productive regime, with maximum productivity in winter.
Another feature that makes this study area particularly relevant is that it has different types of fronts.
On the one hand, there are two strong persistent fronts, the Gulf Stream and the shelf-break front, which are held in place by topography and atmospheric circulation, and which are both associated with strong and deep-reaching vertical circulations \parencite{flagg_2006, liao_2022}.
But there are also more ephemeral and weaker fronts, created by mesoscale strain that are continuously forming at more random locations \parencite{drushka_2019, sanchez-rios_2020}.
We use satellite data of \as{chl} and \as{sst}, and extend on the approach of \textcite{liu_2016}, to distinguish between persistent and ephemeral fronts.
We evaluate the impact of both types of fronts on \as{chl} on the basis of three indicators, the excess (or deficit) \as{chl} over fronts at the local scale of the front, the surplus \as{chl} attributable to fronts at the scale of regional biomes, and the change in the timing of the \as{chl} spring bloom over fronts.

@@

For \as{chl}, we used the L3 product distributed by ACRI-ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (SeaWIFS, MODIS Aqua \& Terra, MERIS, VIIRS-SNPP \& JPSS1, OLCI-S3A \& S3B) merged and reprocessed, available daily at 4km resolution \parencite{chl}.

For \as{sst}, we used the European Space Agency Sea Surface Temperature Climate Change Initiative analysis product version 2.1 \parencite{merchant_2019, good_2020, sst}, also available daily at 4km resolution over the period 2000--2020.
This product combines data from all available infrared sensors ((A)ATSR, SLSTR, and AVHRR sensors), ensuring good resolution where data are available, unlike other \as{sst} products which also include microwave and in-situ measurements, resulting in considerable smoothing of the \as{sst} field. %chktex 36
Where \as{sst} data is not available, spatial interpolation is performed to obtain a cloud free product which, at the cost of resolution on finer features, provides complete synoptic coverage of our large study area.
This interpolation tends to provide an underestimate of the detection of fronts, as the \as{sst} field is smoother over cloud-covered areas \parencite{merchant_2019}.
However, the combination of several sensors allows to reduce these areas to a minimum.
Furthermore, we have only considered cloud-free pixels for our analysis, which ensures that cloudy areas are not taken into account in our quantification.

@@

The permanently oligotrophic regime to the south of our study area, characterized by warm waters and low \as{chl} (Fig.~\ref{fig:zone-separation}), is known as the subtropical gyre permanently stratified biome \parencite{sarmiento_2004} or permanent deep \as{chl} maximum biome \citep{bock_2022}.
Their is no clear physical boundary to the northern limit of this permanent subtropical biome, so we have chosen the latitudinal limit of 32°N to delineate it, which roughly corresponds to the \(0.1 mg m^{-3}\) \as{chl} isocontour in annual mean \as{chl}.

North of 32°N, the seasonal subtropical biome is also mainly oligotrophic, with intermediate levels of \as{chl} and temperature, and characterized by slightly increased productivity in winter.
This second biome is known as the subtropical gyre seasonally stratified biome \parencite{sarmiento_2004}.
It is bounded to the north by the meanders of the Gulf Stream jet.
The Gulf Stream jet conveys warm, salty waters poleward along the Florida coast up to Cape Hatteras (35°N), where the jet separates from the continental shelf and meanders essentially zonally.
The north wall of the jet, so called because of its steep temperature gradient, marks the sharp, sinuous and unsteady northern limit of this biome.

To the north of the Gulf Stream is the Slope sea which extends to the shelf-break, with colder and fresher waters \parencite{linder_1998}.
Aligned with the shelf break, a persistent front with an intensified surface jet separates the shelf waters (excluded from this study) from the slope sea.
This highly productive subpolar biome, known as subpolar waters \parencite{sarmiento_2004} and high-chlorophyll-bloom \citep{bock_2022}, is characterized by a strong spring bloom whose onset is tied to the spring stratification of the mixed-layer.

@@

Our front detection method aims to sort the domain into pixels that are located (or not) over fronts, at each daily time step.
We follow the approach of \textcite{liu_2016} that builds upon the well established Cayula-Cornillon algorithm \parencite{cayula_1992, belkin_2009}, and identify pixels as belonging to fronts when the region in the vicinity of the pixel is characterized by high \as{sst} heterogeneity.
More precisely, on each pixel and for each time step, we compute an Heterogeneity Index~(\as{hi}) which quantifies the heterogeneity of the \as{sst} in a square window comprising a few pixels~(\(7 \times 7\)) centered on the pixel of interest.
Thus \as{hi} measures the \as{sst} heterogeneity at spatial scales less than~\qty{30}{\km}.
The \as{hi} is defined as the weighted sum of the skewness~\ab{skew}, standard deviation~\ab{std}, and bimodality~\ab{bimod} of \as{sst} within the window (\(\am{hi} = a \left( b \ab{skew} + c \ab{std} + d \ab{bimod} \right)\), with a, b, c, d constant normalization coefficients).

We have adapted the original formulation of \textcite{liu_2016} with minor modification in the computation of bimodality and in the relative weighing of each component which are described below.
We computed the bimodality as the L2 norm of the difference between the \as{sst} histogram (with bins of 0.1°C) and a Gaussian fit of the histogram.
We chose this method because it was simpler to implement and more robust than that of \textcite{liu_2016} which computed the normalized absolute difference between a polynomial fit of order 5 of the \as{sst} histogram and a Gaussian fit of the histogram, the polynomial fit often being poorly constrained.
We normalized each component by its variance (b, c and d being defined as the inverse of the standard deviation of each component computed over one year), as opposed to \textcite{liu_2016} who normalized the coefficients by their annual maximum.
This avoided putting too much weight on extrema.
Finally we normalized \as{hi} (coefficient a) such that 95\% of values are below an arbitrary value of 9.5.
For simplicity, the normalization coefficients were computed for year 2007 and used over the entire time series.
We performed sensitivity tests on the different parameters used to compute \as{hi}, namely the number of pixels in the rolling window and the choice of normalization coefficients.
An example of the resulting \as{hi} over a single front is given on Fig.~\ref{fig:zoom}; elevated \as{hi} values are located along the curved \as{sst} gradient to the north east of the window.

At each time step, we sorted the pixels into those belonging to strong fronts (defined as \(\mathrm{\as{hi}} > 10\)), those belonging to weak fronts (defined as \(5 < \mathrm{\as{hi}} < 10\)), and those that do not belong to fronts (when \(\mathrm{\as{hi}} < 5\), and called them ``background'' in the following).
Sorting fronts by range of \as{hi} enabled us to roughly separate the quasi-permanent fronts, which are associated with the strongest \as{sst} gradients (strong fronts), from the more ephemeral ones, associated with weaker gradients (weak fronts).
The choice of the two \as{hi} threshold is somehow arbitrary, but it is supported by the \as{hi} distributions presented below.

\begin{figure*}
  \includegraphics[width=12cm]{front_example.pdf}
  \caption{
    The \as{sst}, \as{chl}, and heterogeneity index (\as{hi}) of a front on the 7 July 2007.
    The plain and dashed contours correspond to \as{hi} values of 5 and 10.
    This front is categorized as weak.
    \as{chl} are elevated inside the front.
  }%
  \label{fig:zoom}
\end{figure*}

@@

Our definition of weak fronts and strong fronts is primarily based on thresholds, derived from the \as{hi} distributions (Fig.~\ref{fig:chl-vs-hi}, red dashed line).
In each biome, the majority of \as{hi} values are below 5, and as mentioned before, we used this threshold to distinguish pixels in the background from those over fronts (when \(\mathrm{\as{hi}} > 5\)).
Secondly, the number of points in the \as{hi} distribution decreases sharply as the \as{hi} value increases above 5, reflecting the fact that fronts with stronger \as{sst} gradients are much less frequent than fronts with weaker gradients, as one would expect.
We used the \as{hi} threshold of 10 to distinguish weak fronts from strong fronts.
This separation is imperfect, as seen for instance in Fig.~\ref{fig:zoom} where a few pixels with \as{hi} values larger than 10 appear at the core of an otherwise weak front.

\begin{figure*}
  \includegraphics[width=12cm]{chl_vs_hi.pdf}
  \caption{
    Normalized distribution of the Heterogeneity Index (\as{hi}, red dashed line) within each biome, and distribution of \as{chl} as a function of \as{hi} (representing front strength), over the full period 2000--2020.
    Shown are the median value of the \as{chl} distributions (solid black line), and 1st and 3rd quartiles (dashed lines).
    Note that 0.5\% of pixels have outstanding large \as{hi} values and are not included here.
  }%
  \label{fig:chl-vs-hi}
\end{figure*}

However, the choice of these two \as{hi} thresholds is also guided by the resulting global spatial climatology of weak and strong fronts (Fig.~\ref{fig:frt-occurrence}).
Weak fronts are abundant, and more or less evenly distributed, over a broad band around and north of the Gulf Stream jet (Fig.~\ref{fig:frt-occurrence}a).
To the south of the Gulf Stream jet, weak fronts are less present, with nevertheless more fronts on the edges of the subtropical gyre (around 28°N) than in its center.
This distribution of weak fronts is consistent with the predominance of mesoscale variability observed along the Gulf Stream system from satellite altimetry \parencite{zhai_2008}, and the injection of eddy kinetic energy north of the Gulf Stream jet by the Gulf Stream extension.
It is thus consistent with the generation process of submesoscale fronts through mesoscale strain.

@@

With the chosen thresholds, the areal proportion of weak fronts tends to increase from South to North, with 7\%, 19\% and 42\% of \as{hi} values comprised between 5 and 10 in the permanent subtropical, seasonal subtropical, and subpolar biome, respectively.
Regarding strong fronts, there are only present in the seasonal subtropical and subpolar biome (as expected from the fact that the Gulf Stream and shelf-break fronts are located within these biomes), where \as{hi} values above 10 account for 6\% and 17\%, respectively.

The fraction of the area occupied by fronts also varies with the seasons, with generally less fronts in summer (Fig.~\ref{fig:ts-climato}d-f).
In the permanent subtropical biome, weak fronts cover on average up to 12\% in spring and drop to 2\% in summer.
In the seasonal subtropical biome, the variation is from 27\% to 13\% for weak fronts, and 8\% to 4\% for strong front.
In the subpolar biome, strong fronts cover between 11\% in summer and 26\% at their peak.
This seasonality is consistent with other estimates that submesoscale activity is greater in winter due to greater mixed-layer thickness compared to summer \parencite{callies_2015}, and also consistent with the most recent modelling results by \textcite{dong_2020} in another Western Boundary system (the Kuroshio Extension) where they show that the strongest submesoscales occur with a lag of about a month after the mixed layer thickness maximum is reached.
An exception to this seasonal pattern is the greater presence of weak fronts in summer in the subpolar biome (50\% in summer versus 35\% in winter), which may be partly explained by the reduction of the \as{hi} over Western Boundary Current fronts in summer, which results in them being counted as weak fronts in summer.
